Explore CNV distributions of reference cells: SCPCP000015

### 2025-05-29 analyses/infercnv-consensus-cell-type/exploratory-notebooks/02_ewings-reference-cnv.Rmd
Updated notebook to include barplot of library proportions, and add some additional text

We have previously observed that the reference cell CNV distributions in pooled references may overlap with tumor CNV distributions.
This notebook aims to explore the extent of this trend for `SCPCP000015`.

@@

```{r, fig.height = 8}
plot_cnv_distributions(infercnv_df, "endo_pooled")
plot_cnv_distributions(infercnv_df, "immune_pooled")
plot_cnv_distributions(infercnv_df, "endo-immune_pooled")
```

@@

Let's look more closely at the `immune` reference to see if any particular libraries might be driving a (minor) trend.
For each inference with the pooled `immune` reference, we'll find the 25\% top reference cells with the highest CNVs.
Do these cells tend to be from a particular library?

First, note how many reference cells come from each library:

```{r}
ref_immune_df <- infercnv_df |>
  dplyr::filter(
    normal_reference == "immune_pooled",
    cell_group == "reference"
  ) |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-", remove = FALSE)

ref_immune_df |>
  dplyr::select(cell_id) |>
  dplyr::distinct() |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-") |>
  dplyr::count(reference_library_id)
```

The table displayed below shows for each inference (`infercnv_library_id`), how many of top 25\% CNV cells in the pooled immune reference are from each library.
For example, of the reference cells with the highest CNVs in the `SCPCL000822` inference, 28 cells came from `SCPCL000822`, 30 cells came from `SCPCL000824`, etc.
Generally, these numbers are consistent across inferences although cells from `SCPCL000824` do seem to be more prevalent in certain samples.
Even so, there doesn't seem to be any major trend here.

```{r}
top25 <- ref_immune_df |>
  dplyr::group_by(infercnv_library_id) |>
  dplyr::slice_max(total_cnv_per_cell, prop = 0.25) |>
  dplyr::ungroup()
```

```{r}
top25 |>
  dplyr::count(infercnv_library_id, reference_library_id) |>
  tidyr::pivot_wider(
    names_from = reference_library_id,
    values_from = n,
    values_fill = 0
  )
```

@@

It does not seem that reference cell CNV distributions are problematically overlapping with tumor CNV distributions now that references were updated to generate these re-processed results.

>>>>

  We have previously observed that the reference cell CNV distributions in pooled references may overlap with tumor CNV distributions.
This notebook aims to explore the extent of this trend for `SCPCP000015`, as well as compare CNV distributions among cell groups for all libraries.

@@

We show a plot faceted by library `inferCNV` was run on for each of the three pooled references explored.

```{r, fig.width = 10, fig.height = 8, message = FALSE}
plot_cnv_distributions(infercnv_df, "endo_pooled")
plot_cnv_distributions(infercnv_df, "immune_pooled")
plot_cnv_distributions(infercnv_df, "endo-immune_pooled")
```

@@

Let's look more closely at the `immune_polled` reference to see if any particular libraries might be driving a (minor) trend.
For each inference with the `immune_pooled` reference, we'll find the 25\% top reference cells with the highest CNVs.
Do these cells tend to be from a particular library?

The barplot below shows for each `inferCNV` inference with the `immune_pooled` reference, the proportion of reference cells contributed by each of the five given libraries.

```{r, fig.width = 8}
top25_df <- infercnv_df |>
  dplyr::filter(
    normal_reference == "immune_pooled",
    cell_group == "reference"
  ) |>
  tidyr::separate(cell_id, into = c("reference_library_id", "barcode"), sep = "-") |>
  # keep only the top 25% of cells
  dplyr::group_by(infercnv_library_id) |>
  dplyr::slice_max(total_cnv_per_cell, prop = 0.25) |>
  dplyr::ungroup()

ggplot(top25_df) +
  aes(x = infercnv_library_id, fill = reference_library_id) +
  geom_bar(position = "fill") +
  labs(
    x = "Library that inferCNV was run on",
    y = "Proportion of cells in the top 25% CNV\nfrom the given reference library"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

We see a generally consistent trend across inferences; cells in the top 25\% are similarly distributed across libraries that contributed to the reference. 
No particular library seems to be biasing higher CNV values in the reference cells.

@@

It does not seem that reference cell CNV distributions are problematically overlapping with tumor CNV distributions now that references were updated to generate these re-processed results.
Overall, the `endo_pooled` reference appears to give the most distinct CNV distributions between reference and tumor cells.
